---
title: "Gene set enrichment analysis"
subtitle: "RNA-seq edition"
author: "Kim Dill-McFarland, <kadm@uw.edu>"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---
```{r include=FALSE}
knitr::opts_chunk$set(fig.width = 3, fig.height = 3)
```

# Overview
In this workshop, we introduce gene set analysis related to RNA-sequencing. In it, we cover:

* Broad MSigDB gene sets
* hypergeometric enrichment with `clusterProfiler`
* gene set enrichment analysis (GSEA) with `fgsea`

We will do all of our work in [RStudio](https://www.rstudio.com/). RStudio is an integrated development and analysis environment for R that brings a number of conveniences over using R in a terminal or other editing environments.

During the workshop, we will build an R script together, which will be posted as 'live_notes' after the workshop [here](https://github.com/hawn-lab/workshops_UW_Seattle/tree/master/2021.07.14_GSEA).

# Prior to the workshop
Please [install R](http://www.r-project.org), [RStudio](https://www.rstudio.com/products/rstudio/download/), and the following packages. See the [setup instructions](notes/introR_setup.html) for more details.

```{r eval=FALSE}
library(tidyverse)
library(org.Hs.eg.db)
library(msigdbr)
library(fgsea)
```

# Additional resources
## Groups

* [Rladies Seattle](https://www.meetup.com/rladies-seattle/) Not just for ladies! A pro-actively inclusive R community with both in-person and online workshops, hangouts, etc.
* [TidyTuesday](https://github.com/rfordatascience/tidytuesday) A weekly plotting challenge
* [R code club](https://www.riffomonas.org/code_club/) Dr. Pat Schloss opened his lab's coding club to remote participation. 
* [Seattle useR Group](https://www.meetup.com/Seattle-useR/)

## Online

* [R cheatsheets](https://www.rstudio.com/resources/cheatsheets/) also available in RStudio under Help > Cheatsheets
* [The Carpentries](https://carpentries.org/)

# R session

```{r}
sessionInfo()
```

***